# ***Econometrics Project***

### Resilience of Secondary Education in Ukraine During Wartime

Hovorov, Pechenenko, Horetska

---
title: "Resilience of Secondary Education in Ukraine
During Wartime"
output: html_document
---

```{r}
library(data.table)
library(ggplot2)
library(patchwork)
library(lmtest)
library(car)
library(sandwich)
library(pROC)
library(tidyverse)
```

```{r}
load_year <- function(path, year) {
  dt <- fread(path, 
            encoding  = "UTF-8", 
            fill      = TRUE,         
            sep       = ";",
            na.strings = c("", "null", "NULL", "н/д"))
  

  col_map <- list(
    "2022" = list(
      math_ball  = "Block3Ball",      
      math_b100  = "Block3Ball100",
      ukr_ball   = "Block1Ball",
      hist_ball  = "Block2Ball"
    ),
    "new" = list(
      math_ball  = "MathBlockBall",
      math_b100  = "MathBlockBall100",
      ukr_ball   = "UkrBlockBall",
      hist_ball  = "HistBlockBall"
    )
  )
  
  fmt <- if (year == 2022) "2022" else "new"
  m   <- col_map[[fmt]]
  
  
  result <- dt[, .(
    Year      = year,
    Math_raw  = as.numeric(get(m$math_ball)),
    Math_b100 = as.numeric(get(m$math_b100)),
    Ukr_raw   = as.numeric(get(m$ukr_ball)),
    Hist_raw  = as.numeric(get(m$hist_ball))
  )]
  
  result <- result[!is.na(Math_raw) & Math_raw > 0]
  
  cat(sprintf("Year %d: %d participants (math)\n", year, nrow(result)))
  return(result)
}


full_data <- rbindlist(list(
  load_year("Odata2022File.csv", 2022),
  load_year("Odata2023File.csv", 2023),
  load_year("Odata2024File.csv", 2024),
  load_year("Odata2025File.csv", 2025)
))

full_data[, Math_z   := scale(Math_raw), by = Year]
full_data[, Math_pct := rank(Math_raw, ties.method = "average") / .N * 100, by = Year]
```

```{r}
data <- read.csv("nmt_merged2021-2025.csv")
data$MathBall100 <- as.numeric(gsub(",", ".", data$MathBall100))

data$SexTypeName <- as.factor(data$SexTypeName)
data$TerTypeName <- as.factor(data$TerTypeName)
data$RegName     <- as.factor(data$RegName)
data$PTRegName   <- as.factor(data$PTRegName)


data$performance <- as.numeric(gsub(",", ".", data$MathBall100))

data <- data[data$performance > 0 & !is.na(data$performance), ]
final_df <- na.omit(data[, c("performance",
                          "SexTypeName",
                          "TerTypeName",
                          "RegName",
                          "Year",
                          "PTRegName"
                          )])

data$Year  <- as.numeric(as.character(data$Year))
data$Birth <- as.numeric(as.character(data$Birth))
data$age   <- data$Year - data$Birth


```

### Exploratory Data Analysis

```{r}
year_colors <- c("2022" = "#E74C3C", "2023" = "#2ECC71", 
                 "2024" = "#3498DB", "2025" = "#9B59B6")

p1 <- ggplot(full_data, aes(x = Math_z, 
                             fill = factor(Year), 
                             color = factor(Year))) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = year_colors) +
  scale_color_manual(values = year_colors) +
  coord_cartesian(xlim = c(-3, 4)) +
  labs(
    title    = "Distribution of Z-normalized Scores",
    subtitle = "Z-scores calculated within each year",
    x = "Z-score", y = "Density",
    fill = "Year", color = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top",
        plot.title    = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10))


p2 <- ggplot(full_data, aes(x = factor(Year), y = Math_z,
                             fill = factor(Year), 
                             color = factor(Year))) +
  geom_violin(alpha = 0.3, trim = TRUE) +
  geom_boxplot(width = 0.12, alpha = 0.7, outlier.shape = NA,
               color = "gray20") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  scale_fill_manual(values = year_colors) +
  scale_color_manual(values = year_colors) +
  labs(
    title    = "Violin Plot of Normalized Scores",
    subtitle = "Median = 0 → average participant of the year",
    x = "Year", y = "Z-score"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none",
        plot.title    = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10))

p3 <- ggplot(full_data, aes(x = Math_pct, color = factor(Year))) +
  stat_ecdf(linewidth = 0.9) +
  scale_color_manual(values = year_colors) +
  labs(
    title    = "ECDF — Empirical Cumulative Distribution Function",
    subtitle = "Overlapping lines indicate identical distributions",
    x = "Percentile Rank (%)", y = "Cumulative Proportion",
    color = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right",
        plot.title    = element_text(size = 12, face = "bold"),
        plot.subtitle = element_text(size = 10))


layout <- (p1 + p2) / p3 +
  plot_annotation(
    title    = "NMT Mathematics Distribution Analysis (2022–2025)",
    subtitle = "Year-specific normalization neutralizes different scoring scales",
    theme = theme(
      plot.title    = element_text(size = 15, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray40")
    )
  ) +
  plot_layout(heights = c(1, 1))

print(layout)
```

### Performance Model: OLS Regression

```{r}

data$RegName <- relevel(data$RegName, ref = "Тернопільська область")
```

```{r}
performance_model <- lm(performance ~ data$SexTypeName + data$RegName  + data$PTRegName + data$Year + data$age, data = data)

summary(performance_model)
```

#### **Assumption diagnostics**

```{r}
cat("\n--- Variance Inflation Factors (VIF) ---\n")
print(vif(performance_model))

cat("\n--- Breusch-Pagan Test (H0: homoscedastic) ---\n")
print(bptest(performance_model))

cat("\n--- Durbin-Watson Test (H0: no autocorrelation) ---\n")
print(dwtest(performance_model))

cat("\n--- Kolmogorov-Smirnov Test for Normality of Residuals ---\n")
ks_result <- ks.test(residuals(performance_model), "pnorm",
                     mean = mean(residuals(performance_model)),
                     sd   = sd(residuals(performance_model)))
print(ks_result)

par(mfrow = c(2, 2))
plot(performance_model)
par(mfrow = c(1, 1))
```

### Migration Model: Logistic Regression

```{r}
nmt_logit <- data %>%
  mutate(
    Migrated = as.integer(!is.na(PTRegName) & as.character(PTRegName) != as.character(RegName)),
    
    Female   = as.integer(SexTypeName == "жіноча"),
    Urban    = as.integer(TerTypeName == "місто"),
    
    War     = as.integer(Year >= 2022),
    WarYear = pmax(Year - 2021, 0),
    
    Frontline_ord = case_when(
      RegName %in% c("Луганська область", "Донецька область")                     ~ 4,
      RegName %in% c("Запорізька область", "Херсонська область",
                     "Харківська область", "Миколаївська область")                ~ 3,
      RegName %in% c("Дніпропетровська область", "Сумська область",
                     "Чернігівська область", "Одеська область")                   ~ 2,
      TRUE                                                                         ~ 1
    ),
    
  
    Frontline_coll = case_when(
      Frontline_ord <= 2 ~ Frontline_ord,
      Frontline_ord >= 3 ~ 3L
    )
  ) %>%
  group_by(Year) %>%
  mutate(
    Math_z = scale(MathBall100)[, 1],
    Math_q = ntile(MathBall100, 5)
  ) %>%
  ungroup() %>%
  filter(!is.na(Migrated), !is.na(Math_z), !is.na(Math_q))

```

```{r}
m_nmt_clean <- glm(
  Migrated ~
    Female * factor(Frontline_ord) + 
    Urban +                           
    factor(Math_q) +                  
    age +                             
    factor(Year),                    
  family = binomial(link = "logit"),
  data = nmt_logit
)

AIC(m_nmt_clean)
```

```{r}
summary(m_nmt_clean)
```

#### Assumption diagnostics

```{r}
coeftest(m_nmt_clean, vcov = vcovCL(m_nmt_clean, cluster = ~RegName))


roc_nmt <- roc(nmt_logit$Migrated, fitted(m_nmt_clean))
cat("AUC:", round(auc(roc_nmt), 4), "\n")


m_null <- glm(Migrated ~ 1, family = binomial(), data = nmt_logit)
mcfadden <- 1 - (logLik(m_nmt_clean) / logLik(m_null))
cat("McFadden R²:", round(mcfadden, 4), "\n")



vif(m_nmt_clean)
```

------------------------------------------------------------------------

### Migration Model: Logistic Regression

```{r}
m_full <- glm(
  Migrated ~
    Female * factor(Frontline_ord) +  # replicates M5 core
    Urban +                            # replicates M5
    factor(Math_q) +                   # replaces Math_raw
    age +                              # replaces Birth, cleaner interpretation
    factor(Year),                      # replaces transport/school_size context
  family = binomial(link = "logit"),
  data = nmt_logit
)
summary(m_full)
AIC(m_full)
```

```{r}
coeftest(m_full, vcov = vcovCL(m_final, cluster = ~RegName))
```

```{r}

dt <- read.csv(
  "merged_dataset_logit.csv",
  sep = ",",
  fileEncoding = "UTF-8",
  stringsAsFactors = FALSE
)
```

```{r}

# migration
dt$Migrated <- ifelse(dt$RegName != dt$PTRegName, 1, 0)

# gender
dt$Female <- ifelse(dt$SexTypeName == "жіноча", 1, 0)

# urban
dt$Urban <- ifelse(grepl("місто", dt$TerTypeName), 1, 0)

```

```{r}
front_rank <- c(
  "Донецька область" = 4L,
  "Луганська область" = 4L,
  "Запорізька область" = 4L,
  "Херсонська область" = 4L,

  "Харківська область" = 3L,
  "Миколаївська область" = 3L,
  "Дніпропетровська область" = 3L,
  "Сумська область" = 3L,
  "Чернігівська область" = 3L,

  "Одеська область" = 2L,
  "Полтавська область" = 2L,
  "Київська область" = 2L,
  "м. Київ" = 2L,
  "Житомирська область" = 2L,
  "Черкаська область" = 2L,
  "Кіровоградська область" = 2L,
  "Хмельницька область" = 2L,

  "Львівська область" = 1L,
  "Івано-Франківська область" = 1L,
  "Тернопільська область" = 1L,
  "Рівненська область" = 1L,
  "Волинська область" = 1L,
  "Закарпатська область" = 1L,
  "Чернівецька область" = 1L,
  "Вінницька область" = 1L
)

dt$Frontline_ord <- front_rank[dt$RegName]
```

```{r}
dt$Math_raw <- as.numeric(gsub(",", "", dt$MathBlockBall100))
```

```{r}
math_threshold <- quantile(dt$Math_raw, 0.75, na.rm = TRUE)
dt$High_Math <- ifelse(dt$Math_raw >= math_threshold, 1, 0)
```

```{r}
logit_data <- subset(
  dt,
  !is.na(Migrated) & !is.na(Female) & !is.na(Urban) &
  !is.na(Frontline_ord) & !is.na(Math_raw) &
  !is.na(transport_share) & !is.na(average_school_size) & !is.na(High_Math)
)
```

```{r}
logit_data$Math_q <- ntile(logit_data$Math_raw, 5)

logit_data <- logit_data %>%
  mutate(
    log_transport   = log(transport_share + 0.001),
    log_school_size = log(average_school_size)
  )
```

```{r}
m5_corrected <- glm(
  Migrated ~ Female * factor(Frontline_ord) + Urban +
    factor(Math_q) +
    log_transport + log_school_size,
  family = binomial(link = "logit"),
  data   = logit_data
)

summary(m5_corrected)
```

```{r}
m5_null <- glm(Migrated ~ 1, family = binomial(), data = logit_data)
lrtest(m5_null, m5_corrected)
```
